Statistical analysis of whole-genome bisulfite-sequencing data

### 2014-12-01 Rmarkdown/wgbs_statistical_analysis.Rmd
Re-factor wgbs_statistical_analysis chapter.

- [x] DONE

I set out a statistical framework for analysing bisulfite-sequencing data. I begin by explaining the various levels of stochasticity in a methylC-seq experiment. Following this, I define the mathematical notation that I use throughout my thesis. I describe key variables, common estimators of these and their statistical properties. The ideas here are simple, although when extended to their full generality there are several subtleties that must be kept in mind.

@@

@@

Even with only a single sample, there are several levels of variation to consider in a bisulfite-sequencing experiment. This variation has a hierarchical structure, which is illustrated in __FIGURE__ and can be separated into pre-sequencing and post-sequencing sources of variation.

@@

A methylation locus is a single cytosine, that is, a CpG, CHG or CHH. The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus with respect to the forward strand, e.g. chr1:+:666. I frequently refer to loci by the subscript $i$ rather than by $pos_{i}$. This means that the distance between the $i^{th}$ and $(i + 1)^{th}$ methylation loci varies along the genome and, for a small number of instances, that the $i^{th}$ and $(i + 1)^{th}$ methylation loci are on separate chromosomes. Generally, the number of methylation loci, $N_{loci}$, is not known, although estimates may be made based from the sample or a reference genome. Not knowing the exact value of $N_{loci}$ is no great concern.

The methylation state of a locus can vary within a sample due to cell-to-cell variability of the methylation state. As discussed, a sample in a bisulfite-sequencing experiment contains DNA that is extracted from hundreds or thousands of cells and each cell can have a slightly different methylation profile. Furthermore, within a diploid cell there are two copies of each chromosome, and therefore two copies of each methylation locus, and these two copies can have different methylation states. Therefore, it is also necessary to consider the next level down in the hierarchy; the DNA fragments within the sample.

@@

The above definitions can be extended to individual methylation loci, 1-tuples, to m-tuples. Mathematically, an m-tuple is denoted by a sequence of methylation loci, $(i, i + 1, \ldots, i + m - 1)$.

@@

\subsubsection{$\mathcal{I}$}

@@

The simplest and most commonly used estimators of $M_{i}$ and $U_{i}$ are $m_{i}$ and $u_{i}$, that is, the number of reads that are methylated and unmethylated at the $i^{th}$ locus, subject to some filtering of the reads and read-positions \citep[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg, Hansen:2011gu}). $M_{i}$ and $U_{i}$ may be estimated per-strand or aggregated across strands in the case of palindromic methylation loci. CpG methylation is typically aggregated across strands.

`methtuple` uses this simple method to estimate $MM_{(i, i + 1)}$, $MU_{(i, i + 1)}$, etc., that is, the frequency of methylation patterns at m-tuples.

@@

Most analyses of bisulfite-sequencing data have focused on estimating the average methylation level at individual methylation loci, $B_{i}$. The simplest estimator is $\beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}$, which has been widely used \citep[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}).

Under a Binomial model for the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample, $M_{i, j} | n_{i, j} \eqd Binomial(n_{i, j}, B_{i, j})$, $\beta_{i, j}$ is the maximum likelihood estimator of $B_{i, j}$. It should be noted that this is conditional on the observed sequencing coverage, $n_{i, j}$.

Recently, more sophisticated methods have been proposed to estimate or model the average methylation level. These methods, which are based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citet{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and the smoothing-methods proposed by \cite{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}. The aim of these methods isn't necessarily to estimate $B_{i, j}$ - rather, it is typically a step along the way to identify differential methylation - but estimates of $B_{i, j}$ can be obtained if so desired.

@@

Several papers have proposed the beta-binomial distribution as a natural model "for describing methylation levels of an individual site across replicates" \cite{Dolzhenko:2014bo}. The 'beta' component of the distribution models the underlying methylation level, $B_{i, j}$, while the 'binomial' component models the sampling of reads by sequencing. Another way to think of this is that the 'beta' component models the biological variability of the data, while the 'binomial' component models the sampling variability of sequencing. This separation of biological and technical variability has proven successful in detecting differential gene expression from RNA-seq data. For example, the `edgeR` software \cite{Robinson:2010cw} uses the negative binomial distribution, which can be thought of as a gamma-poisson mixture distribution, to account for both the biological and sampling variability.

@@

\subsection{Distributions}

@@

Restricting our attention to CpGs, __FIGURE__ shows the column-wise summaries of the EPISCOPE data, that is, the genome-wide distributions of $\beta$-values for each sample. What is immediately clear is that these distributions are bimodal. This bimodality is driven by the fact that CpGs in CpG islands are generally unmethylated whereas those outside of CpG islands are generally methylated, which we can see when we stratify CpGs by their CpG island status.

__TODO: Using the EPISCOPE data create the following:__

- [ ] Genome-wide density
  - [ ] Stratified by CGI
- [ ] Differences (or correlations) of $\beta$-values between strands

```{r, eval = FALSE, echo = FALSE}
# Run on one of the unices
setwd("/wehisan/home/allstaff/h/hickey/methylation_m-tuples/analyses/processed_data/EPISCOPE")
library(MethylationTuples)
library(ggplot2)
library(dplyr)
library(tidyr)
thesis_theme <- theme_classic(base_size = 12)
thesis_theme <- theme_bw(base_size = 12)

x <- readRDS('EPISCOPE_1_tuples_strand_collapsed.rds')

cgi <- read.table('~/methylation_m-tuples/CGI/model-based-cpg-islands-hg19.txt', header = TRUE, stringsAsFactors = FALSE)
cgi <- GRanges(seqnames = cgi$chr, IRanges(cgi$start, cgi$end))
cgi_status <- overlapsAny(x, cgi)

@@

cn <- colnames(beta)
beta <- data_frame(Individual = rep(substr(cn, 1, 3), each = nrow(beta)),
           Tissue = rep(substr(cn, 4, 6), each = nrow(beta)),
           beta = as.vector(beta),
           CGI = rep(cgi_status, ncol(beta)))

a <- beta[sample(nrow(beta), 100000), ]

ggplot(aes(x = beta), data = a) + 
  geom_density(aes(linetype = Tissue)) + 
  facet_grid(. ~ Individual) +
  thesis_theme

ggplot(aes(x = beta), data = a) + 
  geom_density(aes(linetype = Tissue)) + 
  facet_grid(CGI ~ Individual) +
  thesis_theme

```

To do the same plots for the row-wise summaries would be rather tedious since there are many, many more rows (loci) than columns (samples) in the matrix of $\beta$-values. However, a few examples are shown in __FIGURE__. These row-wise summaries are often modelled by the beta distribution \cite[e.g.,]{Hebestreit:2013ko, Lacey:2013iy}. The Beta distribution is a flexible 2-parameter distribution on $[0, 1]$. It can be unimodal, "U"-shaped or "J"-shaped, depending on the choice of parameters. The Beta distribution also includes the Uniform and arcsine distributions as special cases (__TODO: better source than wikipedia__).

The analysis of the row-wise distributions leads to the analysis of differential methylation and differential variability, which are discussed in __SECTION__.

@@

Just as we can explore the distribution of DNA methylation levels within a sample or between samples, so too can we explore co-methylation within a sample or between samples. I examine this in detail in Chapter \ref{chap:co-methylation}

@@

\section{Downstream analyses}\label{sec:downstream}

- [ ] DONE

- Don't rehash the details of Chapter WGBS analysis. Rather summarise these questions as statistical problems in the framework I outline in this chapter.
- DMC: Identify $\bar{B_{i}^{(1)}} \neq \bar{B_{i}^{(2)}}$
- DMR: Identify runs of DMCs
- DVC: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
- DVR: Identify runs of DVCs

\subsection{What is the genome-wide methylation level?}

This is a simple question with a somewhat complex answer.

- Average (weighted) $\beta$-values.
- Lister's method

\subsubsection{A criticism of ``identifying methylcytosines''}{sec:criticism_of_methylcytosines}

__TODO: Simplify and shorten__

\citet{Lister:2008bh}, one of the first papers to include WGBS data, introduced a method "to identify the presence of a methylated cytosine" (detailed in the supplementary material). This idea was also used in subsequent publications from the same group \citep{Lister:2009hy, Lister:2011kg}[^Lister_2013] with similarly high profiles. I believe that this definition of a "methylcytosine" is an unnecessary source of confusion, as I will explain.

[^Lister_2013]: It is worth noting that this idea was not used in the analysis of WGBS data from more recent paper from the same group \citep{Lister:2013et}.

It is necessary to use scare quotes to distinguish the term 'methylcytosine', as used by \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}, from the standard definition meaning a 5-methylcytosine. A 'methylcytosine' is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}. Thus, the idea of identifying 'methylcytosines' is to determine whether each cytosine in the reference genome displays any evidence of methylation in the sample.

The exact procedure is not mathematically described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}. The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy}. \citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}. Because the written descriptions are unclear, and no code is provided to implement the idea, I am unable to determine exactly what was done. What follows is my interpretation based on what is described in \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}.

The idea is to test the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to 'error', where the 'error' is a combination of the estimated sequencing error and the estimated bisulfite-converstion error. Effectively, \citeauthor{Lister:2008bh} are testing the null hypothesis that $\beta_{i, j} = 0$ against the one-side alternative, $\beta_{i, j} > 0$. Any site where the null hypothesis is rejected is declared a 'methylcytosine'. For each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'. Then, apply a false discovery rate correction to the resulting $P_{i}$ and declare all cytosines with a FDR-adjusted $P_{i}$ less than some nominal value[^nominal_fdr] to be "methylcytosines".

[^nominal_fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$. I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure \citep{Benjamini:1995ws}, although this is not explicitly stated.

The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}. It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.

This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A. thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation). This  affects the false discovery rate correction since the number of tests is far greater if all cytosines are simultaneously corrected compared to a separate correction for each context.

Now to why I don't think this is a very useful procedure. Firstly, this isn't actually estimating the number of methylcytosines in the sample; it is estimating the number of "methylcytosines", which are cytosines in the reference genome that appear to be methylated in at least some subset of the sample. These are two different things. If you want to estimate the number of methylcytosines then a better estimator is based on $\frac{1}{N_{loci}} \sum_{i = 1}^{N_{loci}} \beta_{i}$. A simple example makes this clear.

Suppose we have ten cytosines in our reference sequence, each with a true methylation level of $B_{i} = 0.4$ in our sample, which contains multiple cells of known ploidy. Then the true number of methylcytosines is $0.4$ multiplied by the number of cells in our sample multiplied by the ploidy. We can estimate this by computing the average $\beta_{i}$ (an estimate of the average $B_{i}$) and multiplying it by an estimate of the number of molecules and the (known) ploidy. From this system we could easily generate sequencing data where the observed $\beta$-values are all "significantly" non-zero and thus, according to Lister et al.'s procedure, every cytosine is classified as a "methylcytosine".

Secondly, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg} use only those cytosines that are classified as "methylcytosines" in many downstream analyses or use the 0-1 classification rather than the actual $\beta$-values. For example, in \citet[supplementary figure 2a]{Lister:2009hy} they use a Venn diagram to compare the number of "methylcytosines" called in two biological replicates to summarise the concordance between the two biological replicates. A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are "non-zero".

>>>>

I set out a statistical framework for analysing bisulfite-sequencing data. I begin by explaining the various levels of stochasticity and then define the mathematical notation that I use throughout my thesis. I describe key variables, common estimators of these and their statistical properties. The ideas here are simple, although when extended to their full generality there are several subtleties that must be kept in mind.

@@

- Chapter overview
  - One sample
    - Pre-sequencing
    - Post-sequencing
    - Some complications for a single sample
      - What is $\mathcal{I}$?
      - Genetic heterozygosity at a methylation locus
  - Multiple samples
    - Some complications with multiple samples
      - What is $\mathcal{I}$?
  - Parameter estimation
    - Estimating $M$, $U$
    - Estimating $B$
      - Beta-binomial models
      - Smoothing
    - Statistical properties of $\beta$-values
      - Empirical distributions
      * Write a function(s) to do the following and apply it to the EPISCOPE, Lister, Seisenberger and Ziller datasets:
        - Genome-wide distribution of CpG $\beta$-values
        * Kernel density estimates
        * Stratify by CGI status, genic, intragenic, repeats
        * Correlation across strands
        * Effect of sequencing coverage threshold (`min_cov`)
      - What is the genome-wide methylation level?
        * Various figures get bandied about; what's a good summary?
        - A criticism of ``identifying methylcytosines''
      - Correlations
      - Biases
      - Transformations
  - Downstream analyses
    - Differential methylation
      - Experimental design considerations
      - DMCs: Identify $\bar{B_{i}^{(1)}} \neq \bar{B_{i}^{(2)}}$
        - One-group experiments
        - Two-group experiments
        - Multi-group experiments
      - DMRs: Identify runs of DMCs
        - A priori vs. data-driven regions
        - One-group experiments
        - Two-group experiments
        - Multi-group experiments
    - Differential variability
      - DVCs: Identify $\sigma(B_{i}^{(1)}) \neq \sigma(B_{i}^{(2)})$
      - DVRs: Identify runs of DVCs
    - Allele-specific methylation
    - Methylation entropy
    - Epipolymorphism
    - Epiallele detection
    - Integrating DNA methylation data with other genomes data

@@

Even with only a single sample, there are several levels of variation to consider in a bisulfite-sequencing experiment. I find it convenient to separate these into:

1. Pre-sequencing sources of variation
2. Sequencing and post-sequencing sources of variation.

I describe these sources of variation, which are illustrated in Figure \ref{fig:bs-seq_protocol}.

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/bs-seq_protocol.pdf}
\caption{A schematic illustrating several sources of variation in a bisulfite-sequencing experiment. Each sample starts as a population of cells, with potentially different methylation patterns. Each DNA fragment is coloured by its originating cell, although this is unknown in practice. Illustrated are PCR amplication bias (unequal representation of DNA from each cell following PCR amplification), sampling variation (not all DNA fragments are sequenced), alignment errors (not all sequenced fragments are mapped or may be mapped to the wrong location, as is the case for the green fragment), and filtering during methylation calling (not reads are used, the blue read, and reads may have read-positions removed, the black reads).}
\label{fig:bs-seq_protocol}
\end{figure}

@@

A methylation locus is a single cytosine, that is, a CpG, CHG or CHH. The set of these loci is labelled $\mathcal{I} = \{pos_{i}: i = 1, \ldots, N_{loci} \}$, where $pos_{i}$ is the genomic co-ordinates of the $i^{th}$ locus, e.g., chr1:+:666. Generally, the number of methylation loci, $N_{loci}$, is not known, although estimates may be made based from the sample or a reference genome. Not knowing the exact value of $N_{loci}$ is no great concern.

It is sometimes convenient to refer to loci by the subscript $i$ rather than by $pos_{i}$. It is important to remember that the number of base pairs between the $i^{th}$ and $(i + 1)^{th}$ methylation loci varies along the genome. Also, depending on whether the data are stranded, a pair of loci may be on opposite strands. In a small number of instances the $i^{th}$ and $(i + 1)^{th}$ methylation loci are on separate chromosomes, e.g., the last CpG on chromosome 1 and the first CpG on chromosome 2.

The methylation state of a locus can vary within a sample due to cell-to-cell variability of the methylation state. As discussed, a sample in a bisulfite-sequencing experiment contains DNA that is extracted from hundreds or thousands of cells and each cell can have a slightly different methylation profile. Furthermore, within a diploid cell there are two copies of each chromosome, and therefore two copies of each methylation locus, and these two copies can have different methylation states. Therefore, it is also necessary to consider not just the genomic co-ordinates of the locus but from which DNA fragment the methylation state originated.

@@

The above definitions can be extended from individual methylation loci, 1-tuples, to m-tuples. Mathematically, an m-tuple is denoted by a sequence of methylation loci, $(i, i + 1, \ldots, i + m - 1)$.

@@

\subsubsection{What is $\mathcal{I}$?}

@@

It is rare to need direct estimates of $M_{i}$ or $U_{i}$. In order to estimate $M_{i}$ and $U_{i}$, the absolute number of methylated and unmethylated DNA fragments at the $i^{th}$ locus, we would also require an estimate of $|\mathcal{H}_{i}|$, the number of DNA fragments containing the $i^{th}$ locus. Rather, we are generally interested in estimating the proportion of reads that are methylated, $B_{i} = \frac{M_{i}}{M_{i} + U_{i}}$.

@@

The simplest estimator of $B_{i}$ is $\beta_{i} = \frac{m_{i}}{m_{i} + u_{i}}$, which has been widely used \citep[e.g.,]{Cokus:2008fc, Lister:2008bh, Lister:2009hy, Lister:2011kg}). The values of $m_{i}$ and $u_{i}$ are obtained by methylation calling (see Section \ref{sec:methylation_calling}).

Under a Binomial model for the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample, $M_{i, j} | n_{i, j} \eqd Binomial(n_{i, j}, B_{i, j})$, $\beta_{i, j}$ is the maximum likelihood estimator of $B_{i, j}$.

Recently, more sophisticated methods have been proposed to estimate or model the average methylation level. These methods, which are based on $m_{i, j}$ and $u_{i, j}$, include beta-binomial models \citet{Feng:2014iq, Sun:2014fk, Dolzhenko:2014bo}, and smoothing-methods \cite{Hansen:2011gu, Hansen:2012gr, Hebestreit:2013ko}.

@@

Several papers have proposed the beta-binomial distribution since, as noted by \cite{Dolzhenko:2014bo}, it is "a natural model for describing methylation levels of an individual site across replicates" . The 'beta' component of the distribution models the underlying methylation level, $B_{i, j}$, while the 'binomial' component models the sampling of reads by sequencing. Another way to think of this is that the 'beta' component models the biological variability of the data, while the 'binomial' component models the sampling variability of sequencing. This separation of biological and technical variability has proven successful in detecting differential gene expression from RNA-seq data. For example, the `edgeR` software \cite{Robinson:2010cw} uses the negative binomial distribution, which can be thought of as a gamma-poisson mixture distribution, to account for both the biological and sampling variability.

@@

\subsection{Empirical distributions}

- TODO: Lots of figures and discussion. See chapter plan.

@@

\subsubsection{Genome-wide distribution of $\beta$-values}

Restricting our attention to CpGs, Figure \ref{fig:EPISCOPE_beta_distribution} shows the genome-wide distributions of $\beta$-values, that is, the column-wise summaries, for each EPISCOPE sample. What is immediately clear is that these distributions are bimodal. Figure \ref{fig:EPISCOPE_beta_distribution_CGI_stratified} shows that this bimodality is largely driven by the fact that CpGs in CpG islands are mostly unmethylated whereas those outside of CpG islands are generally methylated. Intermediate $\beta$-values, those between $0.2-0.8$, are evidence of heterogeneity.

Both Figures \ref{fig:EPISCOPE_beta_distribution} and  \ref{fig:EPISCOPE_beta_distribution_CGI_stratified} also reveal that the genome-wide distributions of $\beta$-values for a given tissue are very similar between the three individuals. The three adipose-derived tissues, _SA_, _VA_ and _VAT_, have similar genome-wide profiles while the blood-derived tissue, _BUF_, has more intermediate $\beta$-values, potentially indicating more cellular heterogeneity.

```{r EPISCOPE_genome_wide_beta-value_distributions, eval = FALSE, echo = FALSE}
# Run on one of the unices
setwd("/wehisan/home/allstaff/h/hickey/methylation_m-tuples/analyses/processed_data/EPISCOPE")
library(MethylationTuples)
library(ggplot2)
library(dplyr)
library(tidyr)
thesis_theme <- theme_classic(base_size = 20)
thesis_theme <- theme_bw(base_size = 20)

x <- readRDS("EPISCOPE_1_tuples_strand_collapsed.rds")

cgi <- read.table("~/methylation_m-tuples/CGI/model-based-cpg-islands-hg19.txt",
                  header = TRUE, stringsAsFactors = FALSE)
cgi <- GRanges(seqnames = cgi$chr, IRanges(cgi$start, cgi$end))
cgi_status <- overlapsAny(x, cgi)

@@

cn <- colnames(beta)
beta <- data.frame(Individual = rep(substr(cn, 1, 3), each = nrow(beta)),
                   Tissue = rep(substr(cn, 4, 6), each = nrow(beta)),
                   beta = as.vector(beta),
                   CGI = rep(cgi_status, times = ncol(beta)),
                   stringsAsFactors = FALSE)
beta <- tbl_df(beta)

#a <- beta[sample(nrow(beta), 100000), ]

g <- ggplot(aes(x = beta), data = beta) +
  geom_density(size = 1.5) +
  facet_grid(Tissue ~ Individual, labeller = label_both) +
  thesis_theme + scale_colour_brewer(palette = "Dark2") +
  ggtitle(expression(paste("EPISCOPE: Distribution of ", beta, "-values", sep = ""))) +
  xlab(expression(paste(beta, "-values", sep = "")))
ggsave("EPISCOPE_beta_distribution.pdf", g, height = 9, width = 16)

g <- ggplot(aes(x = beta), data = beta) +
  geom_density(aes(colour = CGI), size = 2) +
  facet_grid(Tissue ~ Individual, labeller = label_both) +
  thesis_theme + scale_colour_brewer(palette = "Dark2") +
  ggtitle(expression(paste("EPISCOPE: Distribution of ", beta, "-values", sep = ""))) +
  xlab(expression(paste(beta, "-values", sep = "")))
ggsave("EPISCOPE_beta_distribution_CGI_stratified.pdf", g, height = 9,
       width = 16)
```

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_distribution.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data. Observations have been combined across strands and only CpGs with at least $10\times$ sequencing coverage are included. Most CpGs are highly unmethylated or highly methylated, with the buffy coat samples being the most extreme example of this bimodality. 'Spikes' in the density estimate are due to the discreteness of $\beta$-values. Abbreviations: _BUF_ = buffy coat; _SA_ = subcutaneous adipose; _VA_ = visceral adipose; _VAT_ = adipocytes derived from visceral adipose.}
\label{fig:EPISCOPE_beta_distribution}
\end{figure}

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/EPISCOPE_beta_distribution_CGI_stratified.pdf}
\caption{Kernel density estimates of the genome-wide distribution of CpG $\beta$-values for the EPISCOPE data, stratified by whether the CpG is in a CpG island. Only CpGs with at least $10\times$ sequencing coverage are included. Most CpGs outside of CpG islands are methylated whereas most CpGs within CpG islands are unmethylated. Spikes' in the density estimate are due to the discreteness of $\beta$-values. Abbreviations: _BUF_ = buffy coat; _SA_ = subcutaneous adipose; _VA_ = visceral adipose; _VAT_ = adipocytes derived from visceral adipose; _CGI_ = CpG island.}
\label{fig:EPISCOPE_beta_distribution_CGI_stratified}
\end{figure}

\subsubsection{Strand-specific summaries}

CpG $\beta$-values are often computed by aggregating the $m$ and $u$ counts across the forward and reverse strands. This assumes that the two strands are indeed symmetrically methylated. To investigate this, for each CpG we can compute the $\beta$-values using only the forward-strand reads, $\beta^{+}$, and only the reverse-strand reads, $\beta^{-}$.

```{r EPISCOPE_strand_specific_beta-value_distributions, eval = FALSE, echo = FALSE}
# Run on one of the unices
setwd("/wehisan/home/allstaff/h/hickey/methylation_m-tuples/analyses/processed_data/EPISCOPE")
library(MethylationTuples)
library(ggplot2)
library(dplyr)
library(tidyr)
library(RColorBrewer)

thesis_theme <- theme_classic(base_size = 20)
thesis_theme <- theme_bw(base_size = 20)

y <- readRDS("EPISCOPE_1_tuples.rds")
y_fwd <- y[strand(y) == "+"]
y_rev <- y[strand(y) == "-"]

ol <- findOverlaps(unstrand(y_fwd), shift(unstrand(y_rev), -1L), type = "equal")
beta_fwd <- methLevel(y_fwd, min_cov = 5L)[queryHits(ol), ]
beta_rev <- methLevel(y_rev, min_cov = 5L)[subjectHits(ol), ]

cors <- matrix(unlist(mclapply(1:10, function(min_cov, y_fwd, y_rev, ol) {
  beta_fwd <- methLevel(y_fwd, min_cov = min_cov)[queryHits(ol), ]
  beta_rev <- methLevel(y_rev, min_cov = min_cov)[subjectHits(ol), ]
  sapply(seq_len(ncol(y)), function(i, x, y) {
    cor(x[, i], y[, i], use = "complete.obs")
    }, x = beta_fwd, y = beta_rev)
  }, y_fwd = y_fwd, y_rev = y_rev, ol = ol, mc.cores = 10)), ncol = 10)

beta_pairs <- data.frame(Individual = rep(substr(cn, 1, 3),
                                          each = nrow(beta_fwd)),
                         Tissue = rep(substr(cn, 4, 6), each = nrow(beta_fwd)),
                         beta_fwd = as.vector(beta_fwd),
                         beta_rev = as.vector(beta_rev),
                         stringsAsFactors = FALSE)
beta_pairs <- tbl_df(beta_pairs)

beta_pairs %>% group_by(Individual, Tissue) %>%
  summarise(cor(beta_fwd, beta_rev, use = "complete.obs"))

mypar <- function(a = 1, b = 1, brewer.n = 8, brewer.name = "Dark2",
                  cex.lab = 2, cex.main = 2, cex.axis = 1.5,
                  mar = c(5.1, 5.1, 3.5, 2.1), mgp = c(3, 1, 0), ...) {
    par(mar = mar, mgp = mgp, cex.lab = cex.lab, cex.main = cex.main,
        cex.axis = cex.axis)
    par(mfrow = c(a, b), ...)
    palette(brewer.pal(brewer.n, brewer.name))
}
pdf("EPISCOPE_beta_correlations_across_strand.pdf", width = 16, height = 9)
mypar(4, 3)
sapply(seq_len(ncol(beta_fwd)), function(i, x, y, ns) {
  smoothScatter(x[, i], y[, i],
                xlab = expression(beta["+"]),
                ylab = expression(beta["-"]),
                main = paste0(ns[i], ": cor = ", round(cor(x[, i], y[, i],
                                                        use = "complete.obs"),
                                                    2)))
  }, x = beta_fwd, y = beta_rev, ns = colnames(beta_fwd))
dev.off()
```

\subsection{Interpretation}

In papers about DNA methylation there is often a sentence like the following (given by \cite{Laird:2003kw}): "About 70% of the CpG dinucleotides in the mammalian genome are methylated". In the context of a whole-genome bisulfite-sequencing experiment, this can be interpreted as the expected $\beta$-value of a randomly selected CpG. However, as can be seen from __FIGURE__, the extreme bimodality of the $\beta$-value distribution means that the expected value is not a particularly useful estimate of the methylation level of a particular CpG. To make a useful statement about the methylation level of a particular CpG really requires that we know whether it is within a CpG island.

The "$70\%$" statement can also be interpreted as an estimate of the probability that a cytosine randomly sampled from a haploid copy of a mammalian genome is a methylcytosine. Because in a whole-genome bisulfite-sequencing experiment we measure methylation from a pool of cells, not a single haploid copy of the genome, I do not think it makes sense to describe a particular cytosine in the genome as being a 'methylcytosine'. That is, it only makes sense to describe an observation of a cytosine, $z_{i}$, as being methylated or unmethylated and not to describe a cytosine, $i$, as being a methylated or unmethylated. However, several important whole-genome bisulfite-sequencing papers, \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^Lister_2013], have sought to use the latter definition, which I believe to be an unnecessary source of confusion.

[^Lister_2013]: It is worth noting that this concept was not used in more recent paper from the same group \citep{Lister:2013et}.

\citet{Lister:2008bh} introduced a method "to identify the presence of a methylated cytosine" (detailed in the supplementary material). In the language of these papers, a 'methylcytosine' is a cytosine in the reference genome where "at least s [sic; I believe this should be "a"] subset of the genomes within the sample were methylated" \citep[supplementary material]{Lister:2009hy}. This amounts to testing the hypothesis $H_{0}: \beta_{i, j} = 0$ against the one-sided alternative $H_{1}: \beta_{i, j} > 0$. This can be thought of as testing the null hypothesis that the observed number of methylated reads at the $i^{th}$ cytosine were simply due to 'error', where the 'error' is a combination of the estimated sequencing error and the estimated bisulfite-converstion error.

Although the exact procedure is not particularly well described in any of \citet{Lister:2008bh, Lister:2009hy, Lister:2011kg}[^mathematical_description], nor is any code made available, I believe the method is as follows. For each cytosine compute a P-value, $P_{i} = \sum_{k = m_{i} + 1}^{k = m_{i} + u_{i}} Prob(X = k)$, where $X = Binom(u_{i} + m_{i}, \epsilon))$ and $\epsilon$ is the estimated 'error'. Any site with an FDR-adjusted P-value below a threshold was declared a 'methylcytosine'[^fdr]

[^mathematical_description]: The earliest of these papers, \citet{Lister:2008bh}, includes a short non-mathematical description, while the most detailed description is given in the supplementary material of \citet{Lister:2009hy}. \citet{Lister:2011kg} simply refers to \citet{Lister:2009hy}

[^fdr]: \citet{Lister:2008bh} used an FDR-adjusted P-value cutoff of $0.05$; \citet{Lister:2009hy} used an FDR-adjusted P-value cutoff of $0.01$. I presume the FDR-adjustment to be based on the Benjamini-Hochberg procedure \citep{Benjamini:1995ws}, although this is not explicitly stated. This procedure was performed separately for each methylation context in \citet{Lister:2009hy}, but it is not clear if this is the case for \citet{Lister:2008bh} (a study of _A. thaliani_, which has large amounts of non-CG methylation) or \citet{Lister:2011kg} (a study that includes pluripotent human cell lines that have non-negligible levels of non-CG methylation). This  affects the false discovery rate correction since the number of tests is far greater if all cytosines are simultaneously corrected compared to a separate correction for each context.

The $\epsilon$ are estimated on a per-sample basis, with bisulfite-conversion error rates estimated from the unmethylated chloroplast genome \citep{Lister:2008bh} or from the genome of the lambda phage spike-in control \citep{Lister:2009hy, Lister:2011kg}. It is not clear how the sequencing error rate was estimated, particularly given that the base qualities are not included in the data available from the website.

The proportion of cytosines that are identified by this procedure as 'methylcytosines' is a poor estimator of the probability that a cytosine randomly sampled from a haploid genome is methylated, unless the sample is incredibly homogeneous. For example, suppose we had a sample where the true methylation level of every CpG, $B_{i}$, was $0.2$. Given sufficient sequencing coverage, every CpG in the genome would be declared a 'methylcytosine' when in fact for any haploid copy of the genome only $20\%$ of CpGs would be expected to be methylcytosines.

Furthermore, referring to CpGs as 'methylcytosines' results in a loss of information since two 'methylcytosines' may have very different $\beta$-values. For example, in \citet[supplementary figure 2a]{Lister:2009hy} the authors use a Venn diagram to compare the number of 'methylcytosines' called in two biological replicates to summarise the concordance between the two biological replicates. A far better summary of the biological replicability is to plot the $\beta$-values from each replicate against one another as a scatter plot, as this includes the magnitude of the $\beta$-values and not just whether they are statistically different from zero.

In summary, for bisulfite-sequencing experiments where the DNA for each sample comes from multiple cells I do not think it makes sense, nor is it useful, to refer to individual cytosines, $i$, as being methylated or unmethylated. Instead, it is better to summarise the methylation level at a CpG by a $\beta$-value since this has a natural interpretation as the estimated proportion of haploid genomes in the sample that are methylated at that CpG. Unfortunately, $\beta$-values are not without their own issues, as discussed below.

@@

Just as we can explore the distribution of DNA methylation levels within a sample or between samples, so too can we explore co-methylation within a sample or between samples. I examine this in detail in Chapters \ref{chap:co-methylation_review} and \ref{chap:co-methylation}.

@@

\section{Downstream analyses}\label{sec:downstream_analyses}

As is clear from the length of Chapter \ref{chap:wgbs_bioinformatics_analysis} alone, there is a considerable amount of bioinformatics work before we even consider analysing the data to address the biological question of interest. I call these the 'downstream' analyses, and there are many that might be considered. Using the above statistical framework, I explain the most commonly asked questions.

\subsection{Differential methylation}

A common downstream analysis is the identification of differential methylation, for which many methods have been proposed \citep[e.g.,]{Akalin:2012cmChen:2013eh, Chen:2014jb, Dolzhenko:2014bo, Gokhman:2014cp, Jaffe:2012gx, Lacey:2013iy, Lister:2009hy, Rijlaarsdam:2014hp, Sofer:2013bk, Xie:2014ez, Feng:2014iq, Hebestreit:2013ko, Sun:2014fk, Park:2014ho, Hansen:2012gr}. This includes the identification of differentially methylated cytosines (DMCs) and differentially methylated regions (DMRs).
